---
title: "Data Journalism Lesson 5: Mutating data"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to make new columns of data.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```
# Mutating data

## The basics

One of the most common data analysis techniques is to look at change over time. The most common way of comparing change over time is through percent change. The math behind calculating percent change is very simple, and you should know it off the top of your head. The easy way to remember it is:

`(new - old) / old` 

Or new minus old divided by old. Your new number minus the old number, the result of which is divided by the old number. To do that in R, we can use `dplyr` and `mutate` to calculate new metrics in a new field using existing fields of data. 

So first we'll import the tidyverse so we can read in our data and begin to work with it.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

Now we'll import a common and simple dataset of county  population estimates from the US Census Bureau. Each year, the Census Bureau publishes estimates for states and counties. This one has every county in the US. A common question: who are the winners and losers?  **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/datajournalismfiles/estimates21.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")

```

Your first task is to import the data. For this exercise, you need to simply run this:

```{r mutating-load-data, message=FALSE, warning=FALSE}
counties <- read_csv("http://mattwaite.github.io/datajournalismfiles/estimates21.csv")

ruthless <- soccer %>% mutate(
  gpg = Goals/Games
) %>% 
  arrange(desc(gpg)) %>% slice(1)

snipers <- soccer %>%
  mutate(SoGPct = (SoG/ShAtt)*100
) %>%
  arrange(desc(SoGPct)) %>% slice(1)

maxgames <- soccer %>% summarize(max_games = max(Games))

confsnipers <- soccer %>%
  group_by(Conference) %>%
  summarize(
    total_sog = sum(SoG),
    total_shatt = sum(ShAtt)
  ) %>%
  mutate(SoGPct = (total_sog/total_shatt)*100
) %>%
  arrange(desc(SoGPct)) %>% slice(1)
```
```{r mutating-load-data-exercise, exercise = TRUE}
soccer <- read_csv("http://mattwaite.github.io/sportsdatafiles/ncaa_womens_soccer_totals_2022.csv")
```
```{r mutating-load-data-exercise-solution}
soccer <- read_csv("http://mattwaite.github.io/sportsdatafiles/ncaa_womens_soccer_totals_2022.csv")
```
```{r mutating-load-data-exercise-check}
grade_this_code()
```

Remember, if you want to see the first six rows -- handy to take a peek at your data -- you can use the function `head`.

```{r head-data, exercise=TRUE, exercise.setup = "mutating-load-data"}
head(??????)
```
```{r head-data-solution}
head(soccer)
```
```{r head-data-check}
grade_this_code()
```

### Exercise 1: Calculating percent change 

The code to calculate percent change is pretty simple. Remember, with `summarize`, we used `n()` to count things. With `mutate`, we use very similar syntax to calculate a new value -- a new column of data -- using other values in our dataset. So in this case, we're trying to do (new-old)/old, but we're doing it with fields. 

If we look at what we got when we imported the data, you'll see there's a column called 2021 as the new data, and we'll use 2020 as the old data. So we're looking at one year. Then, to help us, we'll use arrange again to sort it, so we get the fastest growing county over one year. 

NOTE: Pay attention to those backticks. They're the key next to the 1 key on your computer. Those backticks are there because R does not like numbers as column names. When you get a column name that starts with a number or has a space in it, you'll need backticks around it. 

```{r mutate-change, exercise=TRUE, exercise.setup = "mutating-load-data", message=FALSE}
???????? %>% mutate(
  change = (`????` - `????`)/`????`
) 
```
```{r mutate-change-solution, exercise.reveal_solution = FALSE}
counties %>% mutate(
  change = (`2021` - `2020`)/`2020`
) 
```
```{r mutate-change-check}
grade_this_code()
```

Look out on the right and you'll see your change column. But what do you see right away? Do those numbers look like we expect them to? No. 

### Exercise 2: Converting to a percentage.

Those numbers are a decimal -- a really, really small decimal when you see things like e-03, which means that number is to the -10 power, which means move the decimal that many spaces to the left. We want it expressed as a percentage. So let's fix that by multiplying by 100. 

```{r mutate-change2, exercise=TRUE, exercise.setup = "mutating-load-data", message=FALSE}
???????? %>% mutate(
  change = ((`????` - `????`)/`????`)*???
) 
```
```{r mutate-change2-solution, exercise.reveal_solution = FALSE}
counties %>% mutate(
  change = ((`2021` - `2020`)/`2020`)*100
) 
```
```{r mutate-change2-check}
grade_this_code()
```

Better. But ... unless we really care about Autuga County, Alabama, this isn't doing much for us. 


```{r ruthless, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "mutating-load-data", results='asis'}
glue("So in that season, {ruthless$Institution} were the most ruthless team scoring about {format(ruthless$gpg, digits=2)} goals per game.")
```

### Exercise 3: Arranging to answer a question

So who had the most growth last year from the year before? We can find that with arrange, which we learned about in the last tutorial.

```{r mutate-change-percent, exercise=TRUE, exercise.setup = "mutating-load-data", message=FALSE, warning=FALSE}
counties %>% mutate(
  change = ((`2021` - `2020`)/`2020`)*100
) %>%
  arrange(desc(??????))
```
```{r mutate-change-percent-solution, exercise.reveal_solution = FALSE}
soccer %>%
  mutate(SoGPct = (SoG/ShAtt)*100
) %>%
  arrange(desc(SoGPct))
```
```{r mutate-change-percent-check}
grade_this_code()
```

```{r accurate, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "mutating-load-data", results='asis'}
glue("Who is putting the most pressure on the keeper shot after shot?  {snipers$Institution}, who are on goal {format(snipers$SoGPct, digits=1)}% of the time.
     
     A note about this. Note that {snipers$Institution} played in {snipers$Games} matches. The team that played the most played in {maxgames$max_games}. We'll learn how to deal with that in the next tutorial.")
``` 

### Exercise 3: Combining what we know

With this data, we have every team and their stats. But what if we wanted to know which is the most ruthless conference? Here is an example of how you can use what you learned in the last tutorial with what you learned here.

To take a dataset of every team and get each conference, we need to use group by again. And then, to calculate a percentage for the conference, we need to add up the two pieces we need in summarize *before* we mutate a new column. 

We're going to sum the shots on goal and the shot attempts. Then we're going to create a shots on goal percentage -- let's use the shots on goal percentage column name from the previous exercise. 

```{r mutate-change-percent-arrange, exercise=TRUE, exercise.setup = "mutating-load-data", message=FALSE, warning=FALSE}
soccer %>%
  group_by(NOT THE TEAM BUT THE...) %>%
  summarize(
    total_sog = sum(???),
    total_shatt = sum(?????)
  ) %>%
  mutate(NEWCOLUMNNAME = (LITTLETHINGINSUMMARIZE/BIGTHINGINSUMMARIZE)*???
) %>%
  arrange(desc(NEWCOLUMNNAME))
```
```{r mutate-change-percent-arrange-solution, exercise.reveal_solution = FALSE, warning=FALSE}
soccer %>%
  group_by(Conference) %>%
  summarize(
    total_sog = sum(SoG),
    total_shatt = sum(ShAtt)
  ) %>%
  mutate(SoGPct = (total_sog/total_shatt)*100
) %>%
  arrange(desc(SoGPct))
```
```{r mutate-change-percent-arrange-check}
grade_this_code()
```

```{r conf, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "mutating-load-data", results='asis'}
glue("What conference do you not want to be a keeper?  The {confsnipers$Conference} Conference, who are on goal {format(confsnipers$SoGPct, digits=1)}% of the time.")
``` 
