---
title: "Data Journalism Lesson 4: Aggregates"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to take lots of little things and total them up into bigger things.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(glue)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

In this lesson, you will learn about aggregates - a fundamental concept in data analysis. By the end of this tutorial, you'll understand how to group data together and perform calculations on those groups using R and the dplyr package. You'll practice grouping data by one or more variables, counting occurrences within groups, and arranging results to answer questions about your dataset. These skills are essential for summarizing large datasets and uncovering patterns in your data.

## The Basics

One of the most basic bits of data analysis is just simply taking a lot of things -- thousands of things -- and putting them together somehow. If you have a dataset of every police call in a city, you can find out how many times police are called for someone breaking into a house. If you have a dataset of every pass thrown in college football you can find out which quarterbacks threw for the most yards *and* which receivers caught the most passes. 

So how do you put things together? 

First, we need to load our libraries. 

Run this.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

**From now on, if that's not the first line of your notebook, you're probably doing it wrong.**

## Loading data

The first thing we need to do is get some data to work with. We do that by reading it in. In our case, we're going to read data from a csv file -- a comma-separated values file.

The CSV file we're going to read is a list of all calls to UNLPD from 2018 up to 2024.

You can download the data here if you want to use it in your own notebook. **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/datajournalismfiles/crimelogs.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")

```

So step 2, after setting up our libraries, is most often going to be importing data. In order to analyze data, we need data, so it stands to reason that this would be something we'd do very early.

For this exercise, you need to simply run this:

```{r load-data, message=FALSE, warning=FALSE}
calls <- read_csv("https://mattwaite.github.io/datajournalismfiles/crimelogs.csv")

ncalls <- nrow(calls)

totalcalls <- calls |>
  group_by(incident_code) |>
  summarise(
    count = n()
  ) |>
  arrange(desc(count)) |> slice(1)

```
```{r load-data-exercise, exercise = TRUE}
calls <- read_csv("https://mattwaite.github.io/datajournalismfiles/crimelogs.csv")
```
```{r load-data-exercise-solution}
calls <- read_csv("https://mattwaite.github.io/datajournalismfiles/crimelogs.csv")
```
```{r load-data-exercise-check}
grade_this_code()
```

### Exercise 1: Using head to see your data 

Now we can inspect the data we imported. What does it look like? What's in it? What do we have to work with? 

To do that, we use `head()` and **put the name of the variable we created above between the parenthesis** to show the headers and **the first six rows of data**. 

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(____)
```
```{r head-data-solution, exercise.reveal_solution = FALSE}
head(calls)
```
```{r head-data-check}
grade_this_code()
```
<div id="head-data-hint">
**Hint:** The thing you need is to the left of a <- in a block above.
</div>

Let's look at this. As you can see by the data, we have six calls, **which is what we expect** from `head()`. But notice the first row -- the headers. We have things like `incident_code` and `reported` for what type of call it is and the date and time it was reported. 

```{r numberplayers, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("What if we wanted to know what the most common call type to UNLPD is? To do that by hand, we'd have to take each of the {ncalls} records and sort them into piles. Drunks over here. Potheads over here. We'd put them in groups and then count them.")
```

`dplyr` has a **group by** function in it that does just this. A massive amount of data analysis involves grouping like things together at some point. So it's a good place to start.

**The secret to writing code is that much of it is a pattern.** The building blocks largely stay the same. With the tidyverse, this is *especially* true.

To accomplish our goal, we start with the name of the data we created before. The same name we used in `head()` and `glipse()`. We'll take that data and we'll introduce a new operator: `|>`. The best way to read that operator, in my opinion, is to interpret that as "and then do this." So the start of our pattern is `data |>` which you can read as `data and now we'll do this`.

The this, in our case, is `group_by()`. Like `head()` and `glimpse()`, **we need to put something in the parenthesis**. But unlike `head()` and `glimpse()`, this time the something comes from that first row of head. We are grouping data by one of the pieces of the data -- a column name. *If we're trying to group by incident type*, which field looks like it holds the incident name? Look where you ran head above.

### Exercise 2: Group by and count

First, we're going to group our data by the column containing the names of incidents. 

After we group our data together by the thing we want to group it by, we need to count how many things are in each group. We do that first by saying we want to summarize our data (a count is a part of a summary). To get a summary, we have to tell it what we want to summarize. So in this case, we want a count. To get that, let's create a thing called total and set it equal to n(), which is `dplyr`s way of counting something. 

Here's the pattern. **You fill in where there are blanks with what needs to be there.** What you fill in are the two hints from above.

```{r group-by, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
____ |>
  group_by(____) |>
  summarise(
    total = n()
  ) 
```
```{r group-by-solution, exercise.reveal_solution = FALSE}
calls |>
  group_by(incident_code) |>
  summarise(
    total = n()
  )
```
```{r group-by-check}
grade_this_code()
```
<div id="group-by-hint">
**Hint:** The name of your data is the same as what you used in `head()` and `glimpse()`. Then, look carefully at the first row of the output of head. What you need is in bold at the top. Which column has the names of the types of calls in them?
</div>
In this case, we wanted to group together incident types, signified by the field name `incident_code`. After we group the data, we need to count them up. In dplyr, we use `summarize` [which can do more than just count things](https://dplyr.tidyverse.org/reference/summarise.html). Inside the parentheses in summarize, we set up the summaries we want. In this case, we just want a count of the types: `total = n(),` says create a new field, called `total` and set it equal to `n()`, which might look weird, but it's common in stats. The number of things in a dataset? Statisticians call it n. There are n number of call types in this dataset. So `n()` is a function that counts the number of things there are. 

### Exercise 3: Arranging data

And when we run that, we get a list of positions with a count next to them. But it's not in any order. So we'll add another And Then Do This |> and use `arrange`. Arrange does what you think it does -- it arranges data in order. By default, it's in ascending order -- smallest to largest. But if we want to know the type with the most calls, we need to sort it in descending order. In arrange, we need to tell it what we are arranging by. **If you have group_by, then whatever you are arranging by will *always* come from inside summarize.** The pattern looks like this:

```{r group-by-2, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
calls |>
  group_by(____) |>
  summarise(
    total = n()
  ) |> 
  arrange(desc(____))
```
```{r group-by-2-solution, exercise.reveal_solution = FALSE}
calls |>
  group_by(incident_code) |>
  summarise(
    total = n()
  ) |> 
  arrange(desc(total))
```
```{r group-by-2-check}
grade_this_code()
```
<div id="group-by-2-hint">
**Hint:** The only thing new here is the arrange, and what goes in arrange comes from summarize, and it isn't the n() parts.
</div>
What is the most common police call on campus? If you look at the row with the largest total, you get your answer. 

### Exercise 4: Grouping by more than one thing 

We can, if we want, group by more than one thing. Which location has the most of a single call type? To do that, we can group by the type and location in the data:

```{r group-by-3, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
calls |>
  group_by(____, ____) |>
  summarise(
    total = n()
  ) |> 
  arrange(desc(____))
```
```{r group-by-3-solution, exercise.reveal_solution = FALSE}
calls |>
  group_by(incident_code, location) |>
  summarise(
    total = n()
  ) |> 
  arrange(desc(total))
```
```{r group-by-3-check}
grade_this_code()
```
<div id="group-by-3-hint">
**Hint:** If you're reading this, you're going too fast. I told you what to put in group_by in the sentence above this code window. I will do that a lot this semester. Get used to reading carefully. Hints are EVERYWHERE.
</div>

You can learn a bit about students by looking at these simple counts.

## The Recap

Throughout this lesson, you've learned how to use aggregates to analyze data more effectively. You've practiced grouping data using group_by(), summarizing groups with summarise(), and arranging results with arrange(). These powerful functions allow you to quickly answer questions about your data, such as finding the most common types of incidents or identifying locations with the highest occurrences of specific events. Remember, the ability to aggregate data is a crucial skill in data journalism, enabling you to distill large datasets into meaningful insights. As you continue your journey in data analysis, you'll find these techniques invaluable for uncovering stories hidden within your data.

## Terms to Know

- Aggregate: A summary or total created by combining individual data points into groups.
- `group_by()`: A dplyr function used to group data by one or more variables.
- `summarise()`: A dplyr function used to create summary statistics for grouped data.
- `n()`: A function used within summarise() to count the number of rows in each group.
- `arrange()`: A dplyr function used to sort data based on specified columns.
- `desc()`: A function used within arrange() to sort data in descending order.
- Pipe operator (`|>`): An operator used to chain together multiple operations in R, often read as "and then."
